\texttt{treats}: a modular \texttt{R} package for simulating trees and traits.

### 2023-06-07 paper/treats_paper.tex
Update treats_paper.tex
up paper

% 2961w - 30p

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}
\usepackage{pdflscape}
\usepackage{fullpage}
\usepackage{url}
\usepackage{epsfig}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{float}
\usepackage[osf]{mathpazo}
\restylefloat{table}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\newcommand{\treats}{\texttt{treats} }

\begin{document}

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: \treats package.

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf \treats: a modular \texttt{R} package for simulating trees and traits.} 
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$}\\
% , Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
% $^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
% $^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\
}

\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

% 3000-4000 words

\newpage
\begin{abstract}

    \begin{enumerate}
        \item Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
        Simulated data can be used to generate null, base line or neutral models.
        These models can be used either in comparison to observed data to estimate the mechanisms that generated the data.
        Or they can be used to explore, understand and develop theoretical advances by proposing toy models.

        \item In evolutionary biology, such simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
        These evolutionary processes can then be nearly infinitely modified to include much more complex processes to affect the simulations such as traits co-evolution, competition mechanisms or mass extinction events.

        \item Here I present the \treats package, a modular \texttt{R} package for trees and traits simulations.
        This package is based on a simple birth death algorithm from which all core steps can easily be modified by users (hence the modularity).
        It also provides a tidy interface through the \treats object, allowing users to easily run reproducible simulations.

        \item Here I present the core algorithms and their integrated modularity and provide an example of how to use the \treats package when studying disparity, the diversity of traits.
        The \treats package also comes with an extend manual regularly updated following users' questions or suggestions which comes as an extended supplementary material to this paper.

    \end{enumerate}

\end{abstract}

\noindent (Keywords: trees, traits, simulations, birth-death, null-models, ecology, evolution, disparity)\\

\vspace{1.5in}

\newpage

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

\section{Introduction}

% Simulations in biology
Comparing biological patterns is one of the key ways to understand mechanisms in evolutionary biology.
This lead to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology \citep{felsensteinPCM,pennell2013review}.
As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context (i.e. descent with modification) to understand biological mechanisms or concepts \citep{harmon2019book}.
These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see \citealt{bausman2018neutral} for distinctions or common misconceptions) suggesting different processes or mechanisms.
% These types phylogenetic comparative methods analysis of traits through time are routinely done in palaeontology (disparity; \citealt{guillerme2020disparities}) or in ecology (functional diversity; \citealt{mammola2021concepts}).
For example different traits distribution for species with different diets \citep{deepak2023diet},
% different habitats https://onlinelibrary.wiley.com/doi/full/10.1111/jbi.12939
or by comparing some observed pattern to one simulated under null or base conditions\citep{miller2022alternating}.
In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g. mass extinction allowing the surviving species to acquire new morphologies), collect some data to test this mechanism (e.g. some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g. a null model where the traits evolve randomly regardless of an extinction event)\citep{puttick2020complex}.
For such an approach, we need statistical and softwares solutions to simulate trees and data to generate these specific null models.

% What already exists
In practice, these evolutionary simulations can be done relatively easily on computers using a birth-death process \cite{feller1939birthdeath,stadler2010birthdeath,diversitree}
A birth-death process is a continuous time Markov process that had been routinely implemented in \texttt{R} \citep{R} to simulate biological phylogenies (e.g. \citealt{ape}, \citealt{diversitree}).
These algorithm to generate phylogenetic trees can be coupled with other Markov processes to also generate traits, for example using a Brownian Motion process (BM; \citealt{cavalli1967BM}) or an Ornstein Uhlenbeck  (OU; \citealt{lande1976OU}; see \citealt{cooper2016cautionary} for a distinction between both).
In \texttt{R}, this can be done with several already well used and well documented packages.
For example if you want to simulate diversity through time, you can use \texttt{TreeSim} \citep{treesim} to simulate diversity under a set of specific parameters (e.g. speciation and extinction) with some events disrupting the simulations (e.g. mass extinctions).
One can even improve on generating these patterns using \texttt{FossilSim} \citep{fossilsim} to generate a pattern that will take into account fossilization processes.
Or use \texttt{paleobuddy} \citep{paleobuddy} or \texttt{paleotree} \citep{paleotree} to generate palaeontology specific data.
On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in \texttt{RPANDA} \citep{rpanda} or in \texttt{PETER} \citep{puttick2020complex} where the traits are generated stochasticaly through time (given some process) during the birth-death process.

% The problems with the absence of modularity
Although the packages mentioned above are excellent and routinely used with fast and reliable algorithms and associated documentation, they are all designed for specific tasks and don't allow much modification beyond the input parameters designed by the authors.
For example, \texttt{TreeSim} can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g. simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
Or \texttt{PETER} is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
This absence of modularity has hampered the use of complex and question tailored simulations evolutionary biology, although I acknowledge this was not the primary aim of the authors of the packages mentioned above.
This has led workers to often develop their own tools to answer specific questions (e.g. \citealt{puttick2020complex}).
Therefore, we propose \treats a modular \texttt{R} package to simulate both diversity and disparity through time.

\section{Description}
% Very briefly how does it work
\treats is based on the eponymous \treats function that allows to simulate a phylogeny and some trait(s) simultaneously.
The base birth-death algorithm ``grows'' a phylogenetic tree and generates traits for each node and tips in the following manner:

\begin{enumerate}
    \item Generating branch length;
    \item Selecting a lineage among the currently living ones;
    \item Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node).
\end{enumerate}

@@

    4. Generating some trait(s) value(s) for the selected lineage (no either a tip or a node).

In \treats, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called \texttt{``modifiers''} or \texttt{``traits''} (Fig. \ref{Fig:workflow}).
The simulation then outputs a tree (of class \texttt{``phylo''} and a associated table of traits - \texttt{``matrix''}) that can be visualised using the \texttt{plot.treats} function.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{../inst/gitbook/treats_structure.pdf} 
\caption{\treats package workflow: @@@.}
\label{Fig:workflow}
\end{figure}
% \texttt{"events"} are an additional category of modular objects discussed later.

\subsection{Simulating diversity}

To simulate a birth-death tree, two arguments are essentially needed: the birth-death parameters (\texttt{bd.params}) the rate of speciation (birth or $\lambda$), the rate of extinction (death or $\mu$ - this parameter can be equal to 0) and \texttt{stop.rule}, a rule for when to stop the growth of the tree (e.g. when reaching 20 species).

% \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa}
% \texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}

% By default, the three first steps are included in a default modifier that reproduces a simple birth-death process given the input birth and death parameters.

% \begin{enumerate}
%     \item Waiting: drawing a random number from an exponential distribution with a species dependent rate of $n \times (\lambda + \mu)$, where $n$ is the number of species currently alive in the process.
%     \item Selection: randomly choosing on of the $n$ taxa currently alive in the process;
%     \item Speciation: drawing a random number between 0 and 1; if that number is smaller than the speciation relative to the turnover ($\lambda / \lambda+\mu$), the lineage goes extinct %TODO: check!!!
%     ; else, the lineage speciates.
% \end{enumerate}

% These are included by default in a \texttt{treats} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:

% \texttt{## Generating the default modifiers}
% \texttt{my_modifiers <- make.modifiers()}
% \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa with the default modifiers}
% \texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}

% The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
% For example, to simulate the same process but hard coding the modifiers we can use the following

% \texttt{## }

% \subsection{Simulating traits (disparity)}

% multivariate stuff \cite{adams2019multivarPCM}

\subsection{Using modular and dynamic modifiers}

% Dynamic ones will be for bd.params = list(speciation = runif)

\subsection{Introducing timed or conditional events in the simulations}

\section{Discussion}

\subsection{Modularity}

\subsection{\treats compared to other packages}

\subsection{Further directions}

\section{Conclusion}
The \treats is modular and nice to use.

Extended manual following recommendations from \cite{cooper2016dark}

\section{Package location}
The \treats package is available on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

\section{Acknowledgments}
Thanks to Mark Puttick and Alex Slavenko for comments on the early stage of the development of this package. Thanks to @@@. This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).

\bibliographystyle{sysbio}
\bibliography{References}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={treats: a modular R package for simulating trees and traits.},
  pdfauthor={Thomas Guillerme\^{}1},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Thomas Guillerme\(^1\)}
\date{2023-06-07}

\begin{document}
\maketitle

\(^1\) School of Biosciences, University of Sheffield, Sheffield, S10
2TN, United Kingdom.
\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}

Running head: \texttt{treats}: TREes And Traits Simulations

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Simulating biological realistic data is an important step to
  understand and investigate biodiversity. Simulated data can be used to
  generate null, base line or neutral models. These can be used either
  in comparison to observed data to estimate the mechanisms that
  generated the data. Or they can be used to explore, understand and
  develop theoretical advances by proposing toy models.
\item
  In evolutionary biology, simulations often involve the need of an
  evolutionary process where descent with modification is at the core of
  how the simulated data is generated. These evolutionary processes can
  then be nearly infinitely modified to include complex processes that
  affect the simulations such as traits co-evolution, competition
  mechanisms or mass extinction events.
\item
  Here I present the \texttt{treats} package, a modular \texttt{R}
  package for trees and traits simulations. This package is based on a
  simple birth death algorithm from which all steps can easily be
  modified by users. It also provides a tidy interface through the
  \texttt{treats} object, allowing users to easily run reproducible
  simulations. It also comes with an extend manual regularly updated
  following users' questions or suggestions.
\end{enumerate}

Keywords: trees, traits, simulations, birth-death, null-models, ecology,
evolution, disparity

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Comparing biological patterns is one of the key ways to understand
mechanisms in evolutionary biology. This lead to the development of
phylogenetic comparative methods as key methodologically driven topic in
ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
Harmon 2013). As indicated in the name, phylogenetic comparative methods
rely on comparing patterns in a phylogenetic context to understand
biological mechanisms or concepts (Harmon 2019). These comparisons can
be done between observed patterns under different conditions or against
null, neutral or baseline models (see Bausman (2018) for distinctions)
suggesting different processes or mechanisms. For example different
traits distribution for species with different diets (Deepak, Gower, and
Cooper 2023), or habitats (Pinto-Ledezma et al. 2017) or by comparing
some observed pattern to one simulated under null or base conditions
(Miller et al. 2022). In theory, workers can follow the research
pipeline of thinking of a specific mechanism (e.g.~mass extinction
allowing the surviving species to acquire new morphologies), collect
some data to test this mechanism (e.g.~some traits of species across and
extinction event) and then compare these patterns to one simulated under
no specific conditions (e.g.~a null model where the traits evolve
randomly regardless of an extinction event Puttick, Guillerme, and Wills
(2020)). For such an approach, we need statistical and softwares
solutions to simulate trees and data to generate these specific null
models.

In practice, these evolutionary simulations can be done relatively
easily on computers using a birth-death process (Feller 1939; Stadler
2010; FitzJohn 2012). A birth-death process is a continuous time Markov
process that had been routinely implemented in \texttt{R} (R Core Team
2023) to simulate biological phylogenies (e.g.~(Paradis and Schliep
2019), (FitzJohn 2012)). These algorithm to generate phylogenetic trees
can be coupled with other Markov processes to also generate traits, for
example using a Brownian Motion process (BM; Cavalli-Sforza and Edwards
(1967)) or an Ornstein Uhlenbeck (OU; Lande (1976); see Cooper et al.
(2016) for a distinction between both). In \texttt{R}, this can be done
with several already well used and well documented packages. For example
if you want to simulate diversity through time, you can use
\texttt{TreeSim} (Stadler 2011) to simulate diversity under a set of
specific parameters (e.g.~speciation and extinction) with some events
disrupting the simulations (e.g.~mass extinctions). One can even improve
on generating these patterns using \texttt{FossilSim} (Barido-Sottani et
al. 2019) to generate a pattern that will take into account
fossilization processes. Or use \texttt{paleobuddy} (Rosario Petrucci,
Januario, and Quental 2022) or \texttt{paleotree} (Bapst 2012) to
generate palaeontology specific data. On the other hand, if you need to
simulate both diversity and traits through time, this can be done with
specific parameters in \texttt{RPANDA} (Morlon et al. 2016) or in
\texttt{PETER} (Puttick, Guillerme, and Wills 2020) where the traits are
generated stochasticaly through time (given some process) during the
birth-death process.

Although the packages mentioned above are excellent and routinely used
with fast and reliable algorithms and associated documentation, they are
all designed for specific tasks and don't allow much modification beyond
the input parameters designed by the authors. For example,
\texttt{TreeSim} can simulate a birth-death tree with some extinction
event but is not designed to simulated one with an extinction event that
leads to the birth-death process to be not diversity dependent anymore
(e.g.~simulating a release in selection pressure after the extinction
process that leads to a different process dominating speciation). Or
\texttt{PETER} is not designed to simulate a complex set of traits (say
three correlated BM traits and two independent OU ones). This absence of
modularity has hampered the use of complex and question tailored
simulations evolutionary biology, although I acknowledge this was not
the primary aim of the authors of the packages mentioned above. This has
led workers to often develop their own tools to answer specific
questions (e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, we
propose \texttt{treats} a modular \texttt{R} package to simulate both
diversity and disparity through time.

\hypertarget{description}{%
\section{Description}\label{description}}

\texttt{treats} is based on the eponymous \texttt{treats} function that
allows to simulate a phylogeny and some trait(s) simultaneously. The
base birth-death algorithm ``grows'' a phylogenetic tree and generates
traits for each node and tips in the following manner:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Generating branch length;
\item
  Selecting a lineage among the currently living ones;
\item
  Choosing whether that lineage goes extinct (becomes a tip) or
  speciates (becomes a node).
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Generating some trait(s) value(s) for the selected lineage (no either
  a tip or a node).
\end{enumerate}

In \texttt{treats}, these three or four steps are implemented as modular
functions that the user can easily change using an internal class of
objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig.
\ref{Fig:workflow}). The simulation then outputs a tree (of class
\texttt{"phylo"} and a associated table of traits - \texttt{"matrix"})
that can be visualised using the \texttt{plot.treats} function. A third
class of object called \texttt{"events"} can be added to the simulations
to modify it under certain conditions (e.g.~simulating a mass
extinction). Each element in the algorithm described above can be
modified by the user using the implemented functions
\texttt{make.bd.params} to set the birth-death parameters,
\texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set
the birth-death algorithm and \texttt{make.events} to eventually add one
or more events.

\begin{figure}
\hypertarget{figure1}{%
\centering
\includegraphics{../inst/gitbook/treats_structure.png}
\caption{Figure 1: \texttt{treats} package workflow: the \texttt{treats}
algorithm generates a tree and traits using inbuilt \texttt{"traits"}
objects that contain the instructions on how to generate the trait data
(e.g.~which process? how many dimensions?); \texttt{"modifiers"} objects
that contains instructions on how to ``grow'' the tree (e.g.~by linking
speciation to trait values, or to the current number of species); and
\texttt{"events"} objects that can modify the tree structure,
\texttt{"modifiers"} or \texttt{"traits"} depending on specific
conditions (e.g.~80\% of species with positive trait values go extinct
after reaching a specific time).}\label{figure1}
}
\end{figure}

\hypertarget{brief-applied-example}{%
\subsection{Brief applied example}\label{brief-applied-example}}

The modularity structure of the \texttt{treats} packages allows users to
design complex simulation scenarios to fit their specific question and
data. To illustrate this we will look whether it is possible to detect
changes in disparity (i.e.~diversity of traits) in a subset of the data
published from Beck and Lee (2014) implemented in Guillerme (2018). This
dataset contains the ordinated traits for 50 mammalian species across
the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Loading the data and packages}
\KeywordTok{library}\NormalTok{(treats)}
\KeywordTok{library}\NormalTok{(dispRity)}
\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}

\CommentTok{\#\# Creating the time slices}
\NormalTok{time\_slices \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee\_mat99, BeckLee\_tree,}
                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
                              \DataTypeTok{model  =} \StringTok{"acctran"}\NormalTok{,}
                              \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{96}\NormalTok{, }\DataTypeTok{to =} \DecValTok{36}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}10}\NormalTok{))}

\CommentTok{\#\# Calculating disparity on the two first dimensions only}
\NormalTok{observed\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(}\KeywordTok{boot.matrix}\NormalTok{(time\_slices),}
                               \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                               \DataTypeTok{dimensions =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
\caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
in the subset of Beck and Lee (2014)'s dataset. For this specific data,
is it possible to detect potential changes of disparity due to a mass
extinction event?}
\end{figure}

Using this example dataset, one might be interested in testing whether
the K-Pg extinction had an effect on disparity through time. But can
such effect be detected in the first place? We can test this by
simulating some datasets with similar properties as the observed data
and measure changes in disparity in these simulated datasets.

\hypertarget{simulating-trees}{%
\subsubsection{Simulating trees}\label{simulating-trees}}

The first and simplest way is to simulate tree topologies that have
similar properties than the observed one. To do so, we need to use some
\textbf{speciation} parameter indicating the rate at which lineages
speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
\textbf{extinction} parameter indicating the rate at which they go
extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
be crudely calculated from a give tree by counting the number of
speciations and extinctions through time (but see better methods,
e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
Schliep (2019)). We also need a stopping rule for when to stop the
simulations (in our case when reaching 30 time units). This will produce
a single random tree using the input parameters (Fig. 3). Note that I
will not discuss the options here in great details. Much more
information can be found in the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Using the birth{-}death parameters from the observed tree}
\NormalTok{my\_bd\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree)}
\CommentTok{\#\# Setting the stopping rule (stop after 30 time units)}
\NormalTok{stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{30}\NormalTok{)}

\CommentTok{\#\# Simulate the tree}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sim\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
\caption{Figure 3: A randomly simulated tree with similar properties as
the observed one. Note that the time here is expressed in arbitrary
units.}
\end{figure}

\hypertarget{simulating-trees-and-traits}{%
\subsubsection{Simulating trees and
traits}\label{simulating-trees-and-traits}}

For our specific question, we will also need to simulate some traits
associated with each node and tip. For simplicity we will simulate a two
dimensional Brownian Motion trait. To do so, we can create a
\texttt{"traits"} object with the function \texttt{make.traits}. This
results in a 2 dimensional trait space for all the simulated species and
their nodes (Fig. 4).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating a trait in 2 dimensions.}
\NormalTok{my\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process, }\DataTypeTok{n =} \DecValTok{2}\NormalTok{)}

\CommentTok{\#\# Simulate the tree and traits}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
\caption{Figure 4: A randomly simulated tree with a 2 dimensional random
Brownian Motion trait. Orange, light blue and dark blue dots
respectively represent nodes, fossils and living species.}
\end{figure}

\hypertarget{simulating-a-trees-and-traits-with-events}{%
\subsubsection{Simulating a trees and traits with
events}\label{simulating-a-trees-and-traits-with-events}}

To answer our question, we also want to simulate an extinction event. To
do so, we can create two different \texttt{"events"} object with the
function \texttt{make.events}. The first one will simulate a random
extinction event after reaching half of the simulations (15 time units)
and then making three quarter (0.75) of the taxa go extinct. The second
one will simulate a random extinction but based on trait values: after
reaching time 20, all the species with positive trait values will go
extinct. Both scenarios thus illustrate two different types of mass
extinction (selective and random Puttick, Guillerme, and Wills (2020)).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating a random mass extinction}
\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
